analysis

### 2020-07-02 analyses/analysis.Rmd
broke down analyses into separate scripts; did some other things

breaks_2012_yday <- breakpoints_yday(castes, 2012, t1 = 0.75, t2 = 0.15)
breaks_2011_yday <- breakpoints_yday(castes, 2011, t1 = 0.75, t2 = 0.15)
breaks_2010_yday <- breakpoints_yday(castes, 2010, t1 = 0.75, t2 = 0.15)

@@

```
### BB and floral turnover ~ elevation
```{r, include=FALSE}
bb_range <- net %>%
  group_by(site, elev.mean, bb.sp) %>%
  summarize(abund = n()) %>%
  group_by(bb.sp) %>%
  mutate(elev.floor = min(elev.mean),
         elev.ceiling = max(elev.mean),
         elev.range = elev.ceiling - elev.floor,
         elev.med = elev.floor + ((elev.ceiling - elev.floor)/2))

@@

### Per-transect floral abund
fl_abund <- survey %>%
  group_by(site, date, dayofyear, gdd.cum, year, elev.class2, 
           elev.mean, break1.gdd, break2.gdd) %>%
  summarize(fl.abund = sum(flower.cover))

### Per-transect floral richness
fl_rich <- survey %>%
  filter(flower.cover > 0) %>%
  group_by(site, date, dayofyear, gdd.cum, year, elev.class2, 
           elev.mean, break1.gdd, break2.gdd) %>%
  summarize(fl.rich = n())

@@

### GAM surface plots of abundance and diversity metrics
#### HGAM elevation x time x year (I would really like someone -- maybe my future self -- to check that I'm doing this right)
##### Floral abundance
```{r eval=FALSE, include=FALSE}
fl_abund_yday <- gamV(fl.abund ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    family = "tw",
                    method = "REML")

fl_abund_yday2 <- gamV(log(fl.abund + 0.001) ~ year +
                    te(yday, elev.mean, by=year, bs= c("cr", "cr"), k=c(10, 10), m=1),
                    data = abund_div,
                    #family = "scat",
                    method = "REML")

fl_abund_yday3 <- gamV(log(fl.abund + 0.001) ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    #family = "scat",
                    method = "REML")

AIC(fl_abund_yday, fl_abund_yday2, fl_abund_yday3)

check.gamViz(fl_abund_yday)
summary(fl_abund_yday)

print(plot(fl_abund_yday, allTerms = TRUE), pages = 1)

check.gamViz(fl_abund_yday2)
summary(fl_abund_yday2)

print(plot(fl_abund_yday2, allTerms = TRUE), pages = 1)

check.gamViz(fl_abund_yday3)
summary(fl_abund_yday3)

print(plot(fl_abund_yday3, allTerms = TRUE), pages = 1)

fl_abund_gdd <- gam(fl.abund ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    family = "tw",
                    method = "REML")

fl_abund_gdd.vis <- getViz(fl_abund_gdd)
check.gamViz(fl_abund_gdd.vis)
summary(fl_abund_gdd.vis)

plot(fl_abund_gdd.vis, allTerms = TRUE) +
  ggtitle("FL abundance (gdd)") +
  xlab("GDD") +
  ylab("Elevation")
```

##### Floral richness -- since year does not seem to change the overall pattern (though perhaps the magnitude), I could use a simpler model, perhaps with year as a random intercept effect; some explanat
```{r eval=FALSE, include=FALSE}
### By year
fl_rich_yday <- gam(fl.rich ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    method = "REML")

fl_rich_yday.vis <- getViz(fl_rich_yday)
check.gamViz(fl_rich_yday.vis)
summary(fl_rich_yday.vis)

plot(fl_rich_yday.vis, allTerms = TRUE) +
  ggtitle("FL richness (gdd)") +
  xlab("Day of year") +
  ylab("Elevation")

### With year as parametric term
fl_rich_yday2 <- gam(fl.rich ~ year +
                    te(yday, elev.mean, bs= "tp", k= 10, m=1),
                    data = filter(abund_div),
                    method = "REML")

fl_rich_yday2.vis <- getViz(fl_rich_yday2)
check.gamViz(fl_rich_yday2.vis)
summary(fl_rich_yday2.vis)

plot(sm(fl_rich_yday2.vis, 1)) + 
  l_fitRaster() + 
  l_fitContour() + 
  l_points(size = 1) +
  ggtitle("FL richness (yday)") +
  xlab("Day of year") +
  ylab("Elevation")

plot(pterm(fl_rich_yday2.vis, 1))

### By year
fl_rich_gdd <- gam(fl.rich ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    method = "REML")

fl_rich_gdd.vis <- getViz(fl_rich_gdd)
check.gamViz(fl_rich_gdd.vis)
summary(fl_rich_gdd.vis)

plot(fl_rich_gdd.vis, allTerms = TRUE) +
  ggtitle("FL richness (gdd)") +
  xlab("GDD") +
  ylab("Elevation")

### With year as parametric term
fl_rich_gdd2 <- gam(fl.rich ~ year +
                    te(gdd.cum, elev.mean, bs= "tp", k= 10, m=1),
                    data = filter(abund_div),
                    method = "REML")

fl_rich_gdd2.vis <- getViz(fl_rich_gdd2)
check.gamViz(fl_rich_gdd2.vis)
summary(fl_rich_gdd2.vis)

plot(sm(fl_rich_gdd2.vis, 1)) + 
  l_fitRaster() + 
  l_fitContour() + 
  l_points(size = 1) +
  ggtitle("FL richness (gdd)") +
  xlab("Day of year") +
  ylab("Elevation")

plot(pterm(fl_rich_gdd2.vis, 1))
```

##### BB abundance
```{r eval=FALSE, include=FALSE}
bb_abund_yday <- gam(bb.abund ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    method = "REML")

bb_abund_yday.vis <- getViz(bb_abund_yday)
check.gamViz(bb_abund_yday.vis)
summary(bb_abund_yday.vis)

plot(bb_abund_yday.vis, allTerms = TRUE) +
  ggtitle("BB abundance (yday)") +
  xlab("Day of year") +
  ylab("Elevation")

bb_abund_gdd <- gam(bb.abund ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    method = "REML")

bb_abund_gdd.vis <- getViz(bb_abund_gdd)
check.gamViz(bb_abund_gdd.vis)
summary(bb_abund_gdd.vis)

plot(bb_abund_gdd.vis, allTerms = TRUE) +
  ggtitle("BB abundance (gdd)") +
  xlab("GDD") +
  ylab("Elevation")
```

##### BB richness
```{r eval=FALSE, include=FALSE}
bb_rich_yday <- gam(bb.rich ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    method = "REML")

bb_rich_yday.vis <- getViz(bb_rich_yday)
check.gamViz(bb_rich_yday.vis)
summary(bb_rich_yday.vis)
plot(sm(bb_rich_yday.vis, 1)) +
  l_fitRaster() +
  l_fitContour() +
  l_points() +
  ggtitle("bb richness 2010 (yday)") +
  xlab("Day of year") +
  ylab("Elevation")
plot(sm(bb_rich_yday.vis, 2)) +
  l_fitRaster() +
  l_fitContour() +
  l_points() +
  ggtitle("bb richness 2011 (yday)") +
  xlab("Day of year") +
  ylab("Elevation")
plot(sm(bb_rich_yday.vis, 3)) +
  l_fitRaster() +
  l_fitContour() +
  l_points() +
  ggtitle("bb richness 2012 (yday)") +
  xlab("Day of year") +
  ylab("Elevation")

bb_rich_gdd <- gam(bb.rich ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = filter(abund_div),
                    method = "REML")

bb_rich_gdd.vis <- getViz(bb_rich_gdd)
check.gamViz(bb_rich_gdd.vis)
summary(bb_rich_gdd.vis)

plot(bb_rich_gdd.vis, allTerms = TRUE) +
  ggtitle("BB richness (gdd)") +
  xlab("Day of year") +
  ylab("Elevation")
```

@@

visweb(web[[1]])
plotweb(web[[1]])

>>>>

breaks_2012_yday <- breakpoints_yday(castes, 2012, t1 = 0.5, t2 = 0.15)
breaks_2011_yday <- breakpoints_yday(castes, 2011, t1 = 0.5, t2 = 0.15)
breaks_2010_yday <- breakpoints_yday(castes, 2010, t1 = 0.5, t2 = 0.15)

@@

### BB and floral turnover ~ elevation
```{r, include=FALSE}
bb_range <- net %>%
  group_by(site, elev.mean, bb.sp) %>%
  summarize(abund = n()) %>%
  group_by(bb.sp) %>%
  mutate(elev.floor = min(elev.mean),
         elev.ceiling = max(elev.mean),
         elev.range = elev.ceiling - elev.floor,
         elev.med = elev.floor + ((elev.ceiling - elev.floor)/2))

@@

### Per-transect floral abund
fl_abund <- survey %>%
  group_by(site, date, dayofyear, gdd.cum, year, elev.class2, 
           elev.mean, break1.gdd, break2.gdd) %>%
  summarize(fl.abund = sum(flower.cover)) %>%
  rename("yday" = "dayofyear")

### Per-transect floral richness
fl_rich <- survey %>%
  filter(flower.cover > 0) %>%
  group_by(site, date, dayofyear, gdd.cum, year, elev.class2, 
           elev.mean, break1.gdd, break2.gdd) %>%
  summarize(fl.rich = n()) %>%
  rename("yday" = "dayofyear")

@@

### GAM surface plots of abundance and diversity metrics
#### HGAM elevation x time x year (I would really like someone -- maybe my future self -- to check that I'm doing this right)
##### Floral abundance
```{r eval=FALSE, include=FALSE}
fl_abund_yday <- gamV(log(fl.abund + 0.001) ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = fl_abund,
                    method = "REML")

check.gamViz(fl_abund_yday)
summary(fl_abund_yday)
print(plot(fl_abund_yday, allTerms = TRUE), pages = 1)

fl_abund_gdd <- gamV(log(fl.abund + 0.001) ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = fl_abund,
                    method = "REML")

check.gamViz(fl_abund_gdd)
summary(fl_abund_gdd)
print(plot(fl_abund_gdd, allTerms = TRUE), pages = 1)
```

##### Floral richness -- since year does not seem to change the overall pattern (though perhaps the magnitude), I could use a simpler model, perhaps with year as a random intercept effect; some explanat
```{r eval=FALSE, include=FALSE}
fl_rich_yday <- gamV(fl.rich ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = fl_rich,
                    method = "REML")

check.gamViz(fl_rich_yday)
summary(fl_rich_yday)
print(plot(fl_rich_yday, allTerms = TRUE), pages = 1)

fl_rich_gdd <- gamV(fl.rich ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = fl_rich,
                    method = "REML")

check.gamViz(fl_rich_gdd)
summary(fl_rich_gdd)
print(plot(fl_rich_gdd, allTerms = TRUE), pages = 1)

##### BB abundance
```{r eval=FALSE, include=FALSE}
bb_abund_yday <- gamV(sqrt(bb.abund) ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    method = "REML")

check.gamViz(bb_abund_yday)
summary(bb_abund_yday))
print(plot(bb_abund_yday, allTerms = TRUE), pages = 1)

bb_abund_gdd <- gamV(sqrt(bb.abund) ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    method = "REML")

check.gamViz(bb_abund_gdd)
summary(bb_abund_gdd)
print(plot(bb_abund_gdd, allTerms = TRUE), pages = 1)
```

##### BB abundance with species as a "by" factor smooth
##### I am using family = "poisson"; this results in somewhat overdispersed models, but with a higher variance explained than when I use quasipoisson or negative binomial distributions, and the latter two distributions seem to introduce skew. I am more concerned with maximizing the explanatory power of my models than I am with having accurate p-values. We'll see what reviewers think, I guess. 
```{r eval=FALSE, include=FALSE}
### Species-level abundance
bb_abund_sp <- net %>%
  group_by(site, date, dayofyear, gdd.cum, year, elev.class2, 
           elev.mean, break1.gdd, break2.gdd, bb.sp) %>%
  summarize(bb.abund = n()) %>%
  rename(yday = dayofyear) %>%
  ungroup() %>%
  mutate(bb.sp = factor(bb.sp)) %>%
  filter(bb.sp %in% c("bss", "hort", "pasc", "prat", "soro", "wurf")) # we only want the common species I think

### 2010
bb_abund_sp_yday_2010_nb <- gamV(bb.abund ~ te(yday, elev.mean, 
                                             by= bb.sp, 
                                             bs= c("tp", "tp"), 
                                             k=c(8, 8), m=1),
                    data = filter(bb_abund_sp, year == "2010"),
                    family = negbin(10),
                    method = "REML")

bb_abund_sp_yday_2010 <- gamV(bb.abund ~ te(yday, elev.mean, 
                                             by= bb.sp, 
                                             bs= c("tp", "tp"), 
                                             k=c(8, 8), m=1),
                    data = filter(bb_abund_sp, year == "2010"),
                    family = "poisson",
                    method = "REML")

check.gamViz(bb_abund_sp_yday_2010)
summary(bb_abund_sp_yday_2010)
print(plot(bb_abund_sp_yday_2010, allTerms = TRUE), pages = 1)

### 2011
bb_abund_sp_yday_2011 <- gamV(bb.abund ~ te(yday, elev.mean, 
                                             by= bb.sp, 
                                             bs= c("tp", "tp"), 
                                             k=c(8, 8), m=1),
                    data = filter(bb_abund_sp, year == "2011"),
                    family = "poisson",
                    method = "REML")

check.gamViz(bb_abund_sp_yday_2011)
summary(bb_abund_sp_yday_2011)
print(plot(bb_abund_sp_yday_2011, allTerms = TRUE), pages = 1)

### 2012
bb_abund_sp_yday_2012 <- gamV(bb.abund ~ te(yday, elev.mean, 
                                             by= bb.sp, 
                                             bs= c("tp", "tp"), 
                                             k=c(8, 8), m=1),
                    data = filter(bb_abund_sp, year == "2012"),
                    family = "poisson",
                    method = "REML")

check.gamViz(bb_abund_sp_yday_2012)
summary(bb_abund_sp_yday_2012)
print(plot(bb_abund_sp_yday_2012, allTerms = TRUE), pages = 1)

### This time, we'll try modeling the interaction of year and bb.sp in one model. bam() + discrete=TRUE + fREML might speed things up. I hope.

bb_abund_sp_yday_pool <- bam(bb.abund ~ year + te(yday, elev.mean, 
                                          by= interaction(bb.sp, year), 
                                          bs= c("tp", "tp"), 
                                          k=c(10, 10), m=1),
                    data = bb_abund_sp,
                    discrete = TRUE,
                    family = "poisson",
                    method = "fREML")

bb_abund_sp_yday_pool_vis <- getViz(bb_abund_sp_yday_pool)
check.gamViz(bb_abund_sp_yday_pool_vis)
summary(bb_abund_sp_yday_pool_vis)
print(plot(bb_abund_sp_yday_pool_vis, allTerms = TRUE), pages = 19)
```

##### FL abundance by k.type
##### Need to go back through to simplify, include only ktypes that are used heavily
```{r}
fl_abund_ktype <- survey %>%
  group_by(site, date, dayofyear, gdd.cum, year, elev.class2, 
           elev.mean, break1.gdd, break2.gdd, k.type.ss) %>%
  summarize(fl.abund = sum(flower.cover)) %>%
  rename("yday" = "dayofyear")

fl_abund_ktype_yday_pool <- bam(log(fl.abund + 0.001) ~ year + te(yday, elev.mean, 
                                          by= interaction(k.type.ss, year), 
                                          bs= c("tp", "tp"), 
                                          k=c(10, 10), m=1),
                    data = fl_abund_ktype,
                    discrete = TRUE,
                    #family = "poisson",
                    method = "fREML")

fl_abund_ktype_yday_pool_vis <- getViz(fl_abund_ktype_yday_pool)
check.gamViz(fl_abund_ktype_yday_pool_vis)
summary(fl_abund_ktype_yday_pool_vis)
print(plot(fl_abund_ktype_yday_pool_vis, allTerms = TRUE), pages = 10)
```

##### BB richness
```{r eval=FALSE, include=FALSE}
bb_rich_yday <- gamV(bb.rich ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    family = "poisson",
                    method = "REML")

check.gamViz(bb_rich_yday)
summary(bb_rich_yday)
print(plot(bb_rich_yday, allTerms = TRUE), pages = 1)

ggplot(abund_div, aes(elev.mean, bb.rich)) +
  geom_point() +
  facet_wrap(~year)

ggplot(abund_div, aes(gdd.cum, bb.rich)) +
  geom_point() +
  facet_wrap(~year)

ggplot(abund_div, aes(year, bb.rich)) +
  geom_boxplot() +
  facet_wrap(~year)

bb_rich_gdd <- gamV(bb.rich ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    family = "quasipoisson",
                    method = "REML")

check.gamViz(bb_rich_gdd)
summary(bb_rich_gdd)
print(plot(bb_rich_gdd, allTerms = TRUE), pages = 1)
```

##### FL.per.BB
```{r}
fl.per.bb_yday <- gamV(log(fl.per.bb) ~ year +
                    te(yday, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    method = "REML")

check.gamViz(fl.per.bb_yday)
summary(fl.per.bb_yday)
print(plot(fl.per.bb_yday, allTerms = TRUE), pages = 1)

fl.per.bb_gdd <- gamV(log(fl.per.bb) ~ year +
                    te(gdd.cum, elev.mean, by=year, bs= c("tp", "tp"), k=c(10, 10), m=1),
                    data = abund_div,
                    method = "REML")

check.gamViz(fl.per.bb_gdd)
summary(fl.per.bb_gdd)
print(plot(fl.per.bb_gdd, allTerms = TRUE), pages = 1)
```

@@
